---
Date: April 19th, 2015
Title: Does a person's gun ownership status depend on his/her having a college degree?
output:
  html_document:
    number_sections: yes
    theme: cerulean
---

<!-- For more info on RMarkdown see http://rmarkdown.rstudio.com/ -->

<!-- Enter the code required to load your data in the space below. The data will be loaded but the line of code won't show up in your write up (echo=FALSE) in order to save space-->
```{r echo=FALSE}
load(url("http://bit.ly/dasi_gss_data"))
```

<!-- In the remainder of the document, add R code chunks as needed -->

### Introduction:

In this project, I want to address the question: "Does the proportion of gun ownership depend on whether or not the group has a college degree?". Broadly, is there an association between inclination to own a gun to the fact that whether or not one has a college degree.

The reason why I picked this question is that I was curious if more educated communities stay away from  violent weapons. 

### Data:

* Data Collection: I have obtained the data from gss data provided by the course at the url:http://bit.ly/dasi_gss_data. I have focussed my study for the year 2012. 

* Cases: The cases or observations are the responses of the survey given to each individual. Various 
  questions were posed to obtain  either their opinion or status. For this study I am interested in   
  the level of education (degree)  and their gun ownership status.
  
* Variables: The analysis here is between two categorical variables: 1.  whether a person has a  
  college degree or not (college) 2. His or Her gun ownership status (gun). We model  the college   
  degree as the explanetory variable and the gun ownership as the response variable. 

* Study : This is a observational study based on the past data that was collected. This is a random 
  sample  collected from 1974 individuals in the year 2012. 
  
* Scope of inference - generalizability: The population of interest are  in two cateogories. 1. With 
  a college degree. 2. Other without a college degree. Since the sample size is less than 10% of  
  each of the categories, and so are independent. The sample size satifies the condition that the 
  proportion times the sample size is greater than 10. So we can generalize the inference with in the 
  confidence interval.  There may be oher factors that are ignored during this study, for example, 
  political affiliation of a person, which migh influence a persons inclication to carry guns. 
  Because of that, it is possible, we have biased results here. 
  
* Scope of inference - causality: No, causal links cannot be established, as this is a observational 
  study and there might be many confounding factors that are not taken into accout.


```{r echo=FALSE}
gss12 = subset(gss, gss$year == "2012")
gss12_college =  subset(gss12, gss12$degree == "Bachelor" | gss12$degree ==  "Graduate")

prop_col_gun = subset(gss12_college, gss12_college$owngun == "Yes")

prop_col_nogun = subset(gss12_college, gss12_college$owngun == "No")


gss12_nocollege =  subset(gss12, gss12$degree != "Bachelor" & gss12$degree !=  "Graduate")
prop_nocol_gun = subset(gss12_nocollege, gss12_nocollege$owngun == "Yes")

prop_nocol_nogun = subset(gss12_nocollege, gss12_nocollege$owngun == "No")


gundata = matrix(c(nrow(prop_col_gun), nrow(prop_col_nogun), nrow(prop_col_gun)/nrow(gss12_college), nrow(prop_nocol_gun), nrow(prop_nocol_nogun),  nrow(prop_nocol_gun)/nrow(gss12_nocollege)), nrow=3)
colnames(gundata) = c('college', 'nocollege' )
rownames(gundata) =  c('owngun', 'donot-owngun', 'proportion' )
```

### Exploratory data analysis:
```{r echo=FALSE}
gundata.table =  as.table(gundata)
gundata.table
barplot(gundata[1:2,], beside=FALSE, xlab = "college degree", ylab="gun ownership")
```

* From the table, we see a slightly higher proportion of gun owners from the group with no college 
  degree. However, this is a very small difference. We have to find out if this difference is due to  
  the difference in the level of college degree or due to chance, using both confidence interval and 
  Hypothesis testing.
  
### Inference:

The question for inference is, how do people with college degree compare with people without a college degree with respect to gun ownership.

* Null Hypothesis: There is no difference in the proportion of gun ownership between people with   
                  college degree and people without college degree. 

* Alternative Hypothesis: The proprtion of gun ownership is different between people with college  
                  degree and people without a college degree. 

We will  compute Confidence Interval and cross check that result with Hypothesis testing.

#### Conditions for CLT for Confidence interval

The point_estimate of difference between the two proportions is:
```{r echo=FALSE}
p_col = nrow(prop_col_gun)/nrow(gss12_college)
p_nocol = nrow(prop_nocol_gun)/nrow(gss12_nocollege)
diff_point_estimate = p_col  - p_nocol
diff_point_estimate
```

* Independence: 
  1. Within group:
    The sample is a random sample of people and is less than 10% of all the 
    people who has  a college degree in 2012 and people who do not.
  2. Between group:
    The sample of people who go to college are independent of people who do not go. 

* sample size
  The below sample sizes for both groups are greater than 10.
```{r echo=FALSE}
p1 = p_col * (nrow(gss12_college)) 
p2 = (1-p_col) * (nrow(gss12_college)) 
p3 = p_nocol * (nrow(gss12_nocollege) )
p4 = (1-p_nocol) * (nrow(gss12_nocollege) ) 
p1
p2
p3
p4
```

The confidence interval is of the form (point_estimate (+/-) Margin or error).
We need to find the margin of error, which is Z_score X Standard Error. 
For a 95% confidence interval Z_score is set to 1.96.
The Standard error is:

```{r echo=FALSE}
se = sqrt((p_col*(1-p_col)/nrow(gss12_college)) + (p_nocol*(1-p_nocol)/nrow(gss12_nocollege)))
se

```

For a 95% confidence interval, the difference in the gun ownership proportion  is:

```{r echo=FALSE}
CI = array(0,2)
CI[1] =  (p_col-p_nocol) - 1.96*se
CI[2] =  (p_col-p_nocol) + 1.96*se
CI

```

Let's now proceed to Hypothesis testing to verify the result.

Null Hypothesis is:
H0: p_col-p_nocol = 0
And the alternative hypothesis:
HA: p_col-p_nocol != 0

We need to get Pooled proportions for the Null value. Pooled proportion for the Null Hypothesis is:

```{r echo=FALSE}
p_hnot = (nrow(prop_col_gun) + nrow(prop_nocol_gun))/ (nrow(gss12_college) + nrow(gss12_nocollege))
p_hnot
```

#### Conditions for Hypothesis Testing

* Independence
  *   Within group
  The sample is a random sample of people and is less than 10% of all the 
  people who go to or not to go college
  *   Between group
  The sample of people who go to college are independent of people who do not go. 

* Sample size
  Using the pooled proportions, we can verify the conditions for the sample sizes for both groups to 
  make sure they are greater than 10.
  
```{r echo=FALSE}
p1 = p_hnot * (nrow(gss12_college)) 
p2 = (1-p_hnot) * (nrow(gss12_college)) 
p3 = p_hnot * (nrow(gss12_nocollege) )
p4 = (1-p_hnot) * (nrow(gss12_nocollege) ) 
p1
p2
p3
p4
se2= sqrt((p_hnot*(1-p_hnot)/nrow(gss12_college))+ (p_hnot*(1-p_hnot)/nrow(gss12_nocollege)))
```
The Standard Error is also calculated from the Pooled proportion and the sample size of the individual groups:

```{r echo=FALSE}
se2= sqrt((p_hnot*(1-p_hnot)/nrow(gss12_college))+ (p_hnot*(1-p_hnot)/nrow(gss12_nocollege)))
se2
```

The sample distribution of the difference between the two proportions is
nearly normal with mean=0, and standard error = 0.02080434. 

```{r echo=FALSE}
x   <- seq(-0.1,0.1, 0.01)
y = dnorm(x, 0, 0.02080434)
plot(x,y, type="l", main="Distribution of the difference in proportions of gun owenership")

```

The p-value is obtained as follows:
point-estimate of the differense is 
```{r echo=FALSE}
diff = p_col - p_nocol 
diff
```


p_value is the probability of seeing the difference of the two proportions given the Null Hypothesis is True.  

First we  calculate the Z score for the point estimate.
Z_score is:
```{r echo=FALSE}
Z_score = (diff - 0)/se2
Z_score
```
p_value corresposding to the Z_score is:
```{r echo=FALSE}
p_value = pnorm(Z_score, lower.tail = FALSE)
p_value
```

We obtained a p_value of 0.664, which is very high compared to the  5% significance. 
Because of that we fail to reject the Null Hypothesis. So there is not enough evidence that the gun ownership proportions differ based on whether a person has a college degree or not.

The same result is also abtained by looking at the confidence interval we obtained: 
(-0.04936721  0.03167900). 
As we see, we have '0' in our confidence interval, which indicates that we cannot infact reject the  Null Hypothesis.

### Conclusion:

In this analysis, we wanted to check if the proportion of gun ownership depend on whether or not the group under study has a college degree or not.  I have  obtained gun ownership data from 2012 gss survey. 
I  grouped the data into two variables based on the whether the group has a college degree or not and obtained a point estimate for the difference. Then, using a 95% confidence interval and Hypothesis testing, I decided that the difference we saw is purely by chance and there is infact no association between a person having a college degree and his or her gun ownership.

For future study, I am interested in bringing in more factors, like political affiliation, and  Opinion of family income to see if those factorsinfluence gun ownership.  




 


